CS378: Final Project - Data Artifacts \\ \url{https://github.com/sarimaleem/cs378_fp}

### 2023-04-27 paper/main.tex
rough draft of paper done

\documentclass[twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{tabularx,ragged2e,booktabs,caption}
\usepackage{multirow}
\usepackage{color}
\usepackage[top=1in,bottom=1in,left=1.2in,right=1.2in]{geometry}
\usepackage{hyperref}
\usepackage[small]{titlesec}
\usepackage{cite}
\usepackage{graphicx}

@@

\date{}

@@

\begin{abstract}
    \todo{write an abstract describing the core motivation and findings from your project}
\end{abstract}

@@

\todo{This section should include three paragraphs. The first paragraph is to
briefly describe task and data. The second paragraph should describe the
results of your analysis, and the third paragraph describing your fix and main
experimental take aways. }

The model that was used was \cite{clark2020electra}

@@

@@

\begin{figure}
  \includegraphics[width=\linewidth]{./ambiguous.jpg}
  \caption{Data Cartogram of 2000 training examples}
  \label{}
\end{figure}

@@

In order to deal with this, we first created a program that finds ambiguous
examples. We found the variance of the training example based on its confidence
of the gold label from a modeled trained on 500 iterations, 1000 iterations,
1500 iterations, 2000 iterations, 2500 iterations, and 3000 iterations. If th
variance was greater than 0.2, and the confidence of the model was in between
0.6 and 0.7, the example was deemed ambiguous. According to \textit{Swayamdipta
et al.}, training solely ambiguous examples leads to to greater performance
than training on the whole dataset, only easy to learn examples, or only hard to
learn examples. Thus it is logical to train on these examples.

We pulled 1655 ambiguous samples from the training data and trained on those examples.

@@

\todo{
Your writeup should address how effective is your fix, how broadly applicable
is your fix, etc. Providing a single number (overall accuracy) is necessary but
not sufficient here. For example, if your change made the model better on
challenging NLI examples, you could try to quantify that on one or more slices
of the data, give examples of predictions that are fixed, or even use model
interpretation techniques to try to support claims about how your improved
model is doing its ``reasoning.'' (You can look at the papers listed above to
get a sense of how to do such fine-grained evaluation).  You should report
results from a baseline approach (your initial trained model) as well as your
``best'' method. If doing your own project, baselines such as majority class,
random, or a linear classifier are important to see. \textbf{Ablations}: If you
tried several things, analyze the contribution from each one. These should be
\emph{minimal} changes to the same system; try running things with just one
aspect different in order to assess how important that aspect is. This part of
the report should be at least one page.}

% \todo{Describe the data you use, including how many examples are in the training, development, and test sets. Please also provide shallow statistics of your data, with at least the vocabulary size and instruction length. It is best to report all the statistics, including counts, in a table. Please describe how you compute statistics that can be computed in different ways (i.e., vocabulary size). Describe how you pre-process the data, including how you treat casing, tokenization, unknown words, and anything else that you did to the raw data before providing it to your model. If you are using a subset of the data, explain why and discuss tradeoffs (you will also need to back them with experiments).}

% \section{Implementation Details (3pt)}

% \todo{Briefly describe the implementation details. No need to copy details already specified in the assignment. Include any hyper-parameters the model has. If there are any optimizations that you introduced, this is the place to describe them. Did you have to do any special optimizations to make the method work in reasonable time/memory? Describe it here.}

% \section{}

 \section{Related Work (5pt)}
\todo{Briefly discuss prior research papers related to your approach. This will
likely some papers in the project description document. How is your approach
different from existing studies?}
% \paragraph{Ablations (15pt)}

% \todo{Please report your ablation studies on the development set. Make sure to ablate each feature set independently to show they contribute to your final result.  If a feature set does not improve performance, please do not include it -- it doesn't help. You can still discuss it, but there is really no point in including it in your final model. Reporting the results must be done in a table, and each result must be discussed in the text.}

% \section{Error Analysis (7pt)}

% \todo{Qualitative analysis of selected failure examples. Show and discuss error examples from your development set. You must identify certain classes of errors and use the examples to illustrate them. This is often best to show in a table.}

\section{Conclusion (5pt)}
\todo{Brief conclusion summarizing findings from both numerical results and
qualitative analysis.}

\section*{(Optional) AI Assistance}

\todo{If you have used any AI toolkit (either for writing assistance or code
assistance) for your final project, please describe it here. }

>>>>

\documentclass[twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{tabularx,ragged2e,booktabs,caption}
\usepackage{multirow}
\usepackage{color}
\usepackage[top=1in,bottom=1in,left=1.2in,right=1.2in]{geometry}
\usepackage{hyperref}
\usepackage[small]{titlesec}
\usepackage{cite}
\usepackage{graphicx}
\usepackage[autostyle]{csquotes}

@@

\date{April 27, 2023}

@@

\begin{abstract}

This paper describes the training and testing of an ELECTRA model for natural
language inference using the Stanford Natural Language Inference Dataset. The
model's performance is benchmarked using contrast sets, and data cartography is
used to improve the model's accuracy. The results show that the base ELECTRA
model performs well on the dataset, and extracting ambiguous examples from data
cartography and training the model on them is an effective method for improving
its performance for some aspects.

\end{abstract}

@@

Natural language inference is the task of determining the truth value of a
hypothesis given a premise. A hypothesis can either be true if it entails from
the premise, false if it contradicts the premise, or neutral if a determination
cannot be made either way. In this paper, train a model based on the Stanford
Natural Language Inference Dataset (SNLI).

We find that the SNLI model has difficulty differenting between neutral labels,
and labels of other classes. We also find that the model often is unable to
contextualize and relate the semantics of different words that have the same
meaning. We created a contrast set in order to test the model, and found that
the performance of the model is significantly degraded in comparison to the
validation set.

In order to reduce ambiguity between words, we decided to train the model on
ambiguous data - data with high variability and medium confidence. We found that
training that although training the model on ambiguous data reduced the models
overall performance, the performance of the model increased on adversarial data,
which we dubbed contrast sets.

@@

The model that was used was the ELECTRA model \cite{clark2020electra}. We are
training  on the stanford NLI dataset \cite{bowman2015large}. The dataset has
570000 training examples and 10000 validation examples. The baseline ELECTRA
modeled was fine tuned to the SNLI training set and was trained with three
epochs. We used the default loss function from the hugging face transformers
library.

For the fix, we took ambiguous data from the first 50000 training examples and
trained on those for 15 epochs.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{./ambiguous.jpg}
  \caption{Data Cartogram of 2000 training examples, orange ambiguous}
  \label{}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{./difficult.jpg}
  \caption{Data Cartogram of 2000 training examples, orange difficult}
  \label{}
\end{figure}

@@

In order to deal with this, we first created a program that finds ambiguous
examples. We found the variance of the training example based on its confidence
of the gold label from a modeled trained on 500 iterations, 1000 iterations,
1500 iterations, 2000 iterations, 2500 iterations, and 3000 iterations. If th
variance was greater than 0.15, and the confidence of the model was in between
0.4 and 0.8, the example was deemed ambiguous. According to \textit{Swayamdipta
et al.}, training solely ambiguous examples leads to to greater performance
than training on the whole dataset, only easy to learn examples, or only hard to
learn examples. Thus it is logical to train on these examples.

We pulled 1655 ambiguous samples from the training data and trained on those
examples. We trained for 15 epochs on ambiguous examples. As a test, we also
trained solely on difficult to learn examples for five epochs, of which there
were 6915, and benchmarked their performance.

@@

We evaluated the new model that was trained on hard to learn examples and
independently trained on ambiguous examples. We found that models trained on
ambiguous examples tended to do better with more difficult data, but tended to
reduce the models overall average. For example, the accuracy of the model
trained on ambiguous data and tested on the contrast set was 80\%, whereas the
model that was not trained on ambiguous data had an accuracy of 77\%.

Based on manual evaluation of the contrast set, there is some evidence that the
model is able to reason better about differentiaing between entailment and
contradictions (given that they have no errors) and in some instances better
predicts neutrals. The following is a good example of this phenomenon:

\begin{quote}
premise: Under a blue sky with white clouds, a child reaches up to touch the propeller of a plane standing parked on a field of grass. \\
hypothesis: A child is not reaching to touch the propeller out of curiousity.  \\
label: neutral \\
predicted: neutral
\end{quote}

Unlike the the baseline model, which does not see the relationship that
\textit{not} has in the sentence (negating curiousity, not touching), the model
that is trained on ambiguous data is able to distinguish examples like these.

% Please add the following required packages to your document preamble:
% \usepackagemultirow
\begin{table}[]
    \caption{Ambiguous Trained Model on Contrast Set}
    \begin{tabular}{|ll|rrr|}
        \hline
        \multicolumn{2}{|l|}{\multirow{2}{*}{}}                           & \multicolumn{3}{l|}{\textbf{Predicted}}                                                              \\ \cline{3-5} 
        \multicolumn{2}{|l|}{}                                            & \multicolumn{1}{l|}{\textit{0}} & \multicolumn{1}{l|}{\textit{1}}  & \multicolumn{1}{l|}{\textit{2}} \\ \hline
        \multicolumn{1}{|l|}{\multirow{3}{*}{\textbf{True}}} & \textit{0} & \multicolumn{1}{r|}{90}         & \multicolumn{1}{r|}{7.692307692} & 0                               \\ \cline{2-5} 
        \multicolumn{1}{|l|}{}                               & \textit{1} & \multicolumn{1}{r|}{10}         & \multicolumn{1}{r|}{61.53846154} & 30.76923077                     \\ \cline{2-5} 
        \multicolumn{1}{|l|}{}                               & \textit{2} & \multicolumn{1}{r|}{0}          & \multicolumn{1}{r|}{7.692307692} & 92.30769231                     \\ \hline
    \end{tabular}
    \caption*{\textit{ 0=Entailment, 1=Neutral, 2=Contradiction }}
\end{table}

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
\begin{table}[]
    \caption{Ambiguous Trained Model on Validation Set}
    \begin{tabular}{|ll|rrr|}
        \hline
        \multicolumn{2}{|l|}{\multirow{2}{*}{}}                           & \multicolumn{3}{l|}{\textbf{Predicted}}                                                             \\ \cline{3-5} 
        \multicolumn{2}{|l|}{}                                            & \multicolumn{1}{l|}{\textit{0}} & \multicolumn{1}{l|}{\textit{1}} & \multicolumn{1}{l|}{\textit{2}} \\ \hline
        \multicolumn{1}{|l|}{\multirow{3}{*}{\textbf{True}}} & \textit{0} & \multicolumn{1}{r|}{83.53}      & \multicolumn{1}{r|}{9.79}       & 7.04                            \\ \cline{2-5} 
        \multicolumn{1}{|l|}{}                               & \textit{1} & \multicolumn{1}{r|}{6.39}       & \multicolumn{1}{r|}{79.75}      & 13.48                           \\ \cline{2-5} 
        \multicolumn{1}{|l|}{}                               & \textit{2} & \multicolumn{1}{r|}{2.70}       & \multicolumn{1}{r|}{7.48}       & 89.871                          \\ \hline
    \end{tabular}
    \caption*{\textit{ 0=Entailment, 1=Neutral, 2=Contradiction }}
\end{table}

However, the model that was trained on ambiguous data seemed to suffer on the
overall benchmark of the evaluation set. The accuracy dropped from 88\% to 84\%
on the validation set.

The most surprising results were the difficult to learn examples. The accuracy
of that model was significantly lower than base model, clocking in an accuracy
of 13\%. Even a model with random guessing could have done better.

Due to time constraints, more ambiguous labels could not be extracted, but it's
certainly possible that if all 500000 training examples are indexed and
ambiguous ones are found, that the model could have better performance than it
currently does. Furthremore, it is clear that training only on diffult to learn
examples degrades model performance significantly.

 \section{Related Work (5pt)}
Zhang et al. \cite{zhang2022cartography} propose another algorithm that uses
data cartography called \enquote{cartography active learning}. However, their
approach differs slightly in that they actively select ambiguous and difficult
examples \textit{during} training and then adjust the input of the training in
real time to train on the ambiguous data. In contrast, our approach finds
ambiguous and difficult data after an initial training period and then trains on
that data.

\section{Conclusion (5pt)}

In this paper, we evaluate an ELECTRA based ML model on the Stanford Natural
Language Inference dataset. We find that the model has difficulty
differentiating between neutral labels and other classes and that it finds at
times spurious correlations between the premise and hypothesis. We also find
that the model suffers significant performance degradation when tested against a
contrast set. To remedy this, we recommend finding ambiguous labels (labels with
high variablity and medium conference) using data cartography, and train the
model on those data points in addition to the training set. Although not a
perfect method, training on ambiguous labels can help the model learn more
difficult correlations.
